---
title: "Rain_TS_EDA"
author: "Halee Staggs, Landon Padgett, Stephen Reagin"
date: "2023-11-20"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    cache       = TRUE,    
    fig.align   = 'center', 
    fig.path    = 'figs/',  
    results     = 'asis',   
    echo        = TRUE,    
    message     = FALSE,     
    strip.white = TRUE,   
    warning     = FALSE)  
```

# Load Packages
```{r}
#install.packages('tidyverse')
library(tidyverse)
#install.packages('forecast')
library(forecast)
#install.packages('fpp3')
library(fpp3)
#install.packages('corrplot')
library(corrplot)
#install.packages('Hmisc')
library(Hmisc)
#install.packages('stats')
library(stats)
```

# Set Working Directory
```{r}
setwd("~/Precipitation_Time_Series")
```

# Load Data
```{r}
sd_rain <- read_csv("sd_1940_2022.csv")
# View(sd_rain)
```

# Set Seed
```{r}
set.seed(8)
```

# Check Dimensions of Data Frame
```{r}
# 30313 rows, 6 features
dim(sd_rain)
```
# Check for Missing Data
## Temp Average has 24084 missing values, but can be imputed from Max and Min
```{r}
# Check for missing data
summary(sd_rain)
```
# Create New Temp Average Variable that averages max and min values
```{r}
sd_rain$TAVG2 <- (sd_rain$TMAX+sd_rain$TMIN) / 2
```

# Verify Summary of Updated Dataframe
```{r}
summary(sd_rain)
```
# Drop the 2 rows with missing data
```{r}
sd_rain <- sd_rain[!is.na(sd_rain$TAVG2), ]
```

# Create List of Desired Variables
```{r}
all.vars <- c('date', 'PRCP', 'TMAX', 'TMIN', 'TAVG2')
num.vars <- c('PRCP', 'TMAX', 'TMIN', 'TAVG2')
rain.vars <- c('date', 'PRCP')
temp.vars <- c('date', 'TMAX')
```


# EDA
## Numerical Relationships: Temp max, temp min, and average are highly correlated. Can probably drop Max and Min for simplicity. 
```{r}
# Check correlation between rain and temperature data
rcorr(as.matrix(sd_rain[ , (names(sd_rain) %in% num.vars)]), type="pearson")

# Table Displays Pearson coefficient, sample size, and p-value
```
## Inspect Date to See Unit of Measurement
```{r}
# Date is a daily measurement
head(sd_rain$date, 40)
```

### Start Date
```{r}
# Start Date
sort(sd_rain$date)[1]
```

### End Date
```{r}
sort(sd_rain$date, decreasing = T)[1]
```

## Time Plots

* Date and Precipitation Time Plot

### Plot shows seasonality but no trend
```{r}
# Configure Plot Settings
prcp.plot <- ggplot(sd_rain, aes(x = date, y = PRCP)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Precipitation by Year', x = 'Years 1940 - 2022', y = 'Precipitation (cm)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

# Save Figure
png(file="prcp.plot.png")
prcp.plot
dev.off()
```
```{r}
# Display plot
prcp.plot
```

* Date and Average Temperaure Time Plot

### Plot shows seasonality and small trend
```{r}
# Configure Plot Settings
temp.plot <- ggplot(sd_rain, aes(x = date, y = TAVG2)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Average Temperature by Year', x = 'Years 1940 - 2022', y = 'Temperature (Celsius)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

# Save Figure
png(file="temp.plot.png")
temp.plot
dev.off()
```
```{r}
# Display plot
temp.plot
```

* Date and Minumum Temperaure Time Plot

### Plot shows seasonality and small trend
```{r}
# Configure Plot Settings
min.plot <- ggplot(sd_rain, aes(x = date, y = TMIN)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Minumum Temperature by Year', x = 'Years 1940 - 2022', y = 'Temperature (Celsius)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

# Save Figure
png(file="min.plot.png")
min.plot
dev.off()
```
```{r}
# Display plot
min.plot
```

* Date and Maximum Temperaure Time Plot

### Plot shows seasonality but no trend
```{r}
# Configure Plot Settings
max.plot <- ggplot(sd_rain, aes(x = date, y = TMAX)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Maximum Temperature by Year', x = 'Years 1940 - 2022', y = 'Temperature (Celsius)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

#Save Figure
png(file="max.plot.png")
max.plot
dev.off()
```
```{r}
# Display plot
max.plot
```

# Time Series Objects 

* Date and Precipitation Data
```{r}
# Create time series
rain.ts <- ts(sd_rain[ , rain.vars])
# Inspect first few values to confirm
head(rain.ts)
```

* Date and Average Temperature Data
```{r}
# Create Time Series
temp.ts <- ts(sd_rain[ , temp.vars])

# Inspect first few values to confirm
head(rain.ts)
```

# Autocorrelation Plots for MA value
```{r}
# Rain ACF
rain.acf <- acf(diff(rain.ts[ , 2]))
rain.acf
```

```{R}
# Temp ACF
temp.acf <- acf(diff(temp.ts[ , 2]))
temp.acf
```

# PACF for AR Value
```{r}
# Rain PACF
rain.pacf <- pacf(diff(rain.ts[ , 2]))
rain.pacf
```

```{r}
# Temp PACF
temp.pacf <- pacf(diff(temp.ts[ , 2]))
temp.pacf
```

# Divide Data into Training and Test Sets for Rain
```{r}
rain.train <- window(rain.ts, start = 1, end = 26293)
rain.train.X <- as.matrix(rain.train[ , 2])
rain.train.y <- rain.train[ , 1]
rain.test <- window(rain.ts, start = 26293)
rain.test.X <- as.matrix(rain.test[ , 2])
rain.test.y <- rain.test[ , 1]
```

# Divide Data into Training and Test Sets for Temperature
```{r}
temp.train <- window(temp.ts, start = 1, end = 26293)
temp.train.X <- as.matrix(temp.train[ , 2])
temp.train.y <- temp.train[ , 1]
temp.test <- window(temp.ts, start = 26293)
temp.test.X <- as.matrix(temp.test[ , 2])
temp.test.y <- temp.test[ , 1]
```


# Train Auto-Arima Models to Start
## Rain auto ARIMA: Random Walk
```{r}
#Auto model to start
rain.auto <- auto.arima(rain.train.y, xreg = rain.train.X)
summary(rain.auto)
```
## Temp auto ARIMA:
```{r}
#Auto model to start
temp.auto <- auto.arima(temp.train.y, xreg = temp.train.X)
summary(temp.auto)
```
