---
title: "Rain_TS_EDA"
author: "Halee Staggs, Landon Padgett, Stephen Reagin"
date: "2023-11-20"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    cache       = TRUE,    
    fig.align   = 'center', 
    fig.path    = 'figs/',  
    results     = 'asis',   
    echo        = TRUE,    
    message     = FALSE,     
    strip.white = TRUE,   
    warning     = FALSE)  
```

# Load Packages
```{r}
#install.packages('tidyverse')
library(tidyverse)
#install.packages('forecast')
library(forecast)
#install.packages('fpp3')
library(fpp3)
#install.packages('corrplot')
library(corrplot)
#install.packages('Hmisc')
library(Hmisc)
#install.packages('stats')
library(stats)
#install.packages('tibbletime')
library(tibbletime)
#install.packages('caret')
library(caret)
```

# Set Working Directory
```{r}
setwd("~/Precipitation_Time_Series")
```

# Load Data
```{r}
sd_rain <- read_csv("sd_1940_2022.csv")
#View(sd_rain)
```

# Set Seed
```{r}
set.seed(8)
```

# Check Dimensions of Data Frame
```{r}
# 30313 rows, 6 features
dim(sd_rain)
```
# Check for Missing Data
## Temp Average has 24084 missing values, but can be imputed from Max and Min
```{r}
# Check for missing data
summary(sd_rain)
```
# Create New Temp Average Variable that averages max and min values
```{r}
sd_rain$TAVG2 <- (sd_rain$TMAX+sd_rain$TMIN) / 2
```

# Create New Variable with Temp Daily Swing
```{r}
sd_rain$TDIFF <- sd_rain$TMAX-sd_rain$TMIN
```

# Verify Summary of Updated Dataframe
```{r}
summary(sd_rain)
```
# Drop the 2 rows with missing data
```{r}
sd_rain <- sd_rain[!is.na(sd_rain$TAVG2), ]
```

# Create List of Desired Variables
```{r}
all.vars <- c('date', 'PRCP', 'TMAX', 'TMIN', 'TAVG2', 'TDIFF')
num.vars <- c('PRCP', 'TMAX', 'TMIN', 'TAVG2', 'TDIFF')
rain.vars <- c('date', 'PRCP')
temp.vars <- c('date', 'TDIFF')
```


# EDA
## Numerical Relationships: Temp max, temp min, and average are highly correlated. Can probably drop Max and Min for simplicity. 
```{r}
# Check correlation between rain and temperature data
rcorr(as.matrix(sd_rain[ , (names(sd_rain) %in% num.vars)]), type="pearson")

# Table Displays Pearson coefficient, sample size, and p-value
```
## Inspect Date to See Unit of Measurement
```{r}
# Date is a daily measurement
head(sd_rain$date, 40)
```

### Start Date
```{r}
# Start Date
sort(sd_rain$date)[1]
```

### End Date
```{r}
sort(sd_rain$date, decreasing = T)[1]
```

## Look at Distributions of Numerical Variables
### There is a decent portion of the sample with large values.
```{r}
# Precipitation BoxPlot
ggplot(sd_rain, aes(x = PRCP)) +
  geom_histogram(bins = 30) +
  labs(x = 'Precipitation (mm)', y = 'Frequency', title = 'Precipitation Distribution') +
  theme_classic()
```
```{r}
# Inspect lower precipitation values
ggplot(sd_rain %>% filter(PRCP < 10), aes(x = PRCP)) +
  geom_histogram(bins = 30) +
  labs(x = 'Precipitation (mm)', y = 'Frequency', title = 'Precipitation Distribution') +
  theme_classic()
```
```{r}
# Maximum Temperature BoxPlot
ggplot(sd_rain, aes(x = TMAX)) +
  geom_histogram(bins = 30) +
  labs(x = 'Temperature (Celsius)', y = 'Frequency', title = 'Maximum Temperature Distribution') +
  theme_classic()
```
```{r}
# Minimum Temperature BoxPlot
ggplot(sd_rain, aes(x = TMIN)) +
  geom_histogram(bins = 30) +
  labs(x = 'Temperature (Celsius)', y = 'Frequency', title = 'Minumum Temperature Distribution') +
  theme_classic()
```
```{r}
# Average Temperature BoxPlot
ggplot(sd_rain, aes(x = TAVG2)) +
  geom_histogram(bins = 30) +
  labs(x = 'Temperature (Celsius)', y = 'Frequency', title = 'Average Temperature Distribution') +
  theme_classic()
```
```{r}
# Temperature Swing BoxPlot
ggplot(sd_rain, aes(x = TDIFF)) +
  geom_histogram(bins = 30) +
  labs(x = 'Temperature (Celsius)', y = 'Frequency', title = 'Temperature Swing Distribution') +
  theme_classic()
```

## Time Plots for Daily Data

* Date and Precipitation Time Plot

### Plot shows seasonality but no trend
```{r}
# Configure Plot Settings
prcp.plot <- ggplot(sd_rain, aes(x = date, y = PRCP)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Precipitation by Year', x = 'Years 1940 - 2022', y = 'Precipitation (cm)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

# Save Figure
png(file="prcp.plot.png")
prcp.plot
dev.off()
```
```{r}
# Display plot
prcp.plot
```

* Date and Average Temperaure Time Plot

### Plot shows seasonality and small trend
```{r}
# Configure Plot Settings
temp.plot <- ggplot(sd_rain, aes(x = date, y = TAVG2)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Average Temperature by Year', x = 'Years 1940 - 2022', y = 'Temperature (Celsius)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

# Save Figure
png(file="temp.plot.png")
temp.plot
dev.off()
```
```{r}
# Display plot
temp.plot
```

* Date and Minumum Temperaure Time Plot

### Plot shows seasonality and small trend
```{r}
# Configure Plot Settings
min.plot <- ggplot(sd_rain, aes(x = date, y = TMIN)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Minumum Temperature by Year', x = 'Years 1940 - 2022', y = 'Temperature (Celsius)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

# Save Figure
png(file="min.plot.png")
min.plot
dev.off()
```
```{r}
# Display plot
min.plot
```

* Date and Maximum Temperaure Time Plot

### Plot shows seasonality but no trend
```{r}
# Configure Plot Settings
max.plot <- ggplot(sd_rain, aes(x = date, y = TMAX)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Maximum Temperature by Year', x = 'Years 1940 - 2022', y = 'Temperature (Celsius)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

#Save Figure
png(file="max.plot.png")
max.plot
dev.off()
```
```{r}
# Display plot
max.plot
```

* Date and Temperaure Swing Time Plot

### Plot shows seasonality but no trend
```{r}
# Configure Plot Settings
diff.plot <- ggplot(sd_rain, aes(x = date, y = TDIFF)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Temperature Difference by Year', x = 'Years 1940 - 2022', y = 'Temperature (Celsius)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

#Save Figure
png(file="diff.plot.png")
diff.plot
dev.off()
```

## Time Plots for Weekly Data
### Singal is weak for this period
```{r}
#Transform dataset into weekly periods with max rainfall
#View(sd_rain_wk)
sd_rain_wk <- as_tbl_time(sd_rain[ , rain.vars], index = date) %>%
collapse_by("week") %>%
group_by(date) %>%
summarise_if(is.numeric, max)
```

* Week and Precipitation Timeplot
```{r}
# Configure Plot Settings
prcp.wk.plot <- ggplot(sd_rain_wk, aes(x = date, y = PRCP)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Precipitation by Week', x = 'Years 1940 - 2022', y = 'Precipitation (mm)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

prcp.wk.plot

#Save Figure
png(file="prcp.wk.plot.png")
diff.plot
dev.off()
```


## Time Plots for Monthly Data
### Monthly period gained a little bit of signal
```{r}
#Transform dataset into monthly periods with max rain
#View(sd_rain_mn)
sd_rain_mn <- as_tbl_time(sd_rain[ , rain.vars], index = date) %>%
collapse_by("month") %>%
group_by(date) %>%
summarise_if(is.numeric, max)
```

* Month and Precipitation
```{r}
# Configure Plot Settings
prcp.mn.plot <- ggplot(sd_rain_mn, aes(x = date, y = PRCP)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Precipitation by Month', x = 'Years 1940 - 2022', y = 'Precipitation (mm)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

prcp.mn.plot

#Save Figure
png(file="prcp.mn.plot.png")
diff.plot
dev.off()
```

* Month and Precipitation for Last 22 Years Only
```{r}
prcp.mn.plot.trim <- ggplot(sd_rain_mn %>% filter(date > '1999-12-31'), aes(x = date, y = PRCP)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Precipitation by Month', x = 'Years 2000 - 2022', y = 'Precipitation (mm)') +
  scale_x_date(date_breaks = "1 year", date_labels = "%m") +
  stat_smooth(color = 'orange', lwd = 3) 

prcp.mn.plot.trim
```
* Month and Temperature Swing for last 22 Years
```{r}
#Transform dataset into monthly periods with temp swing
#View(sd_temp_sn)
sd_temp_mn <- as_tbl_time(sd_rain[ , temp.vars], index = date) %>%
collapse_by("1 month") %>%
group_by(date) %>%
summarise_if(is.numeric, max)
```



```{r}
# Configure Plot Settings
tdiff.mn.plot.trim <- ggplot(sd_temp_mn %>% filter(date > '1999-12-31'), aes(x = date, y = TDIFF)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Largest Temperature Swing by Month', x = 'Years 2000 - 2022', y = 'Temperature (Celsius)') +
  scale_x_date(date_breaks = "1 year", date_labels = "%m") +
  stat_smooth(color = 'orange', lwd = 3) 

tdiff.mn.plot.trim

#Save Figure
#png(file="tdiff.sn.plot.png")
#diff.plot
#dev.off()
```


## Timeplots for Seasonal Data
### Seasonal data has the most signal
```{r}
#Transform dataset into 3-month periods with max rain
View(sd_rain_sn)
sd_rain_sn <- as_tbl_time(sd_rain[ , rain.vars], index = date) %>%
collapse_by("3 month") %>%
group_by(date) %>%
summarise_if(is.numeric, max)
```

* Season and Precipitation
```{r}
# Configure Plot Settings
prcp.sn.plot <- ggplot(sd_rain_sn, aes(x = date, y = PRCP)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Precipitation by Season', x = 'Years 1940 - 2022', y = 'Precipitation (mm)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

prcp.sn.plot

#Save Figure
png(file="prcp.sn.plot.png")
diff.plot
dev.off()
```

* Season and Precipitation for the Past 22 years
```{r}
# Configure Plot Settings
prcp.trim.sn.plot <- ggplot(sd_rain_sn %>% filter(date > 1999-12-31), aes(x = date, y = PRCP)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Precipitation Max by Season', x = 'Years 2000 - 2022', y = 'Precipitation (mm)') +
  #scale_x_date(date_breaks = "1 year") +
  stat_smooth(color = 'orange', lwd = 3) 

prcp.trim.sn.plot

#Save Figure
png(file="prcp.trim.sn.plot.png")
diff.plot
dev.off()
```



* Season and Temp Difference
```{r}
#Transform dataset into 3-month periods with temp swing
#View(sd_temp_sn)
sd_temp_sn <- as_tbl_time(sd_rain[ , temp.vars], index = date) %>%
collapse_by("3 month") %>%
group_by(date) %>%
summarise_if(is.numeric, max)
```

```{r}
# Configure Plot Settings
tdiff.sn.plot <- ggplot(sd_temp_sn, aes(x = date, y = TDIFF)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Largest Temperature Swing by Season', x = 'Years 1940 - 2022', y = 'Temperature (Celsius)') +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

tdiff.sn.plot

#Save Figure
png(file="tdiff.sn.plot.png")
diff.plot
dev.off()
```

```{r}
# Configure Plot Settings
tdiff.trim.sn.plot <- ggplot(sd_temp_sn %>% filter(date > 1999-12-31), aes(x = date, y = TDIFF)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Largest Temperature Swing by Season', x = 'Years 2000 - 2022', y = 'Temperature (Celsius)') +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  stat_smooth(color = 'orange', lwd = 3) 

tdiff.trim.sn.plot

#Save Figure
png(file="tdiff.sn.plot.png")
diff.plot
dev.off()

```

# Time Series Objects with Past 22 Years of Data

* Date and Precipitation Data for Seasons
```{r}
# Create time series
rain.ts.sn <- ts(sd_rain_sn %>% filter(date > '1999-12-31'), frequency = 4)
# Inspect first few values to confirm
head(rain.ts.sn)
```

* Date and Temperature Data for Seasons
```{r}
# Create Time Series
temp.ts.sn <- ts(sd_temp_sn %>% filter(date > '1999-12-31'), frequency = 4)
# Inspect first few values to confirm
head(temp.ts.sn)
```


* Date and Precipitation Data for Months
```{r}
# Create time series
rain.ts.mn <- ts(sd_rain_mn %>% filter(date > '1999-12-31'), frequency = 12)
# Inspect first few values to confirm
head(rain.ts.mn)
```

* Date and Temperature Data for Months
```{r}
# Create Time Series
temp.ts.mn <- ts(sd_temp_mn %>% filter(date > '1999-12-31'), frequency = 12)
# Inspect first few values to confirm
head(temp.ts.mn)
```



# Seasonality Analysis: Values are correlated with seasonal cycles (every 4th lag) and a negative autocorrelation at 1 suggesting swings in the data - for both rain and temperature 
## Autocorrelation Plots for MA value
### Season periodicity
```{r}
# Rain ACF
rain.acf.sn <- acf(diff(rain.ts.sn[ , 2]))
rain.acf.sn
```

```{R}
# Temp ACF
temp.acf.sn <- acf(diff(temp.ts.sn[ , 2]))
temp.acf.sn
```

# PACF for AR Value (value should be 3)
```{r}
# Rain PACF
rain.pacf.sn <- pacf(diff(rain.ts.sn[ , 2]))
rain.pacf.sn
```

```{r}
# Temp PACF
temp.pacf.sn <- pacf(diff(temp.ts.sn[ , 2]))
temp.pacf.sn
```

### Monthly periodicity
```{r}
# Rain ACF
rain.acf.mn <- acf(diff(rain.ts.mn[ , 2]))
rain.acf.mn
```

```{R}
# Temp ACF
temp.acf.mn <- acf(diff(temp.ts.mn[ , 2]))
temp.acf.mn
```

# PACF for AR Value (value should be 3)
```{r}
# Rain PACF
rain.pacf.mn <- pacf(diff(rain.ts.mn[ , 2]))
rain.pacf.mn
```

```{r}
# Temp PACF
temp.pacf.mn <- pacf(diff(temp.ts.mn[ , 2]))
temp.pacf.mn
```



# Data Partition for Seasonal Data
# Divide Data into Training and Test Sets for Rain
```{r}
#rain.train <- window(rain.ts, start = 1, end = 160)
#rain.test <- window(rain.ts, start = 161)
#rain.train.X <- as.matrix(rain.train[ , 2])
#rain.test.X <- as.matrix(rain.test[ , 2])
#rain.train.y <- rain.train[ , 1]
#rain.test.y <- rain.test[ , 1]
```

# Divide Data into Training and Test Sets for Temperature
```{r}
#temp.train <- window(temp.ts, start = 1, end = 160)
#temp.train.X <- as.matrix(temp.train[ , 2])
#temp.train.y <- temp.train[ , 1]
#temp.test <- window(temp.ts, start = 161)
#temp.test.X <- as.matrix(temp.test[ , 2])
#temp.test.y <- temp.test[ , 1]
```


# Data Partition for Monthly Data
```{r}
#View(rain.ts.mn)
rain.train <- window(rain.ts.mn, start = 1, end = c(22,12))
rain.test <- window(rain.ts.mn, start = c(23,1))
rain.train.X <- as.matrix(rain.train[ , 2])
rain.test.X <- as.matrix(rain.test[ , 2])
rain.train.y <- rain.train[ , 1]
rain.test.y <- rain.test[ , 1]
```
```{r}
rain.test
```

# Divide Data into Training and Test Sets for Temperature
```{r}
temp.train <- window(temp.ts.mn, start = 1, end = c(22,12))
temp.train.X <- as.matrix(temp.train[ , 2])
temp.train.y <- temp.train[ , 1]
temp.test <- window(temp.ts.mn, start = c(23,1))
temp.test.X <- as.matrix(temp.test[ , 2])
temp.test.y <- temp.test[ , 1]
```

# Naive Forecast for Monthly Precipiation
```{r}
#View(rain.train)
naive.rain <- naive(rain.train.X, h = 12)
```

```{r}
autoplot(naive.rain)
```


# Seasonal Naive Forecast for Monthly Precipitation
```{r}
snaive.rain <- snaive(rain.train.X, h = 12)
```
```{R}
autoplot(snaive.rain)
```

# Exponential Smoothing Model
```{r}
# First difference at 12-lag for monthly seasonality
rain.diff <- diff(rain.ts.mn, lag = 12)

# Split data based on differenced data
rain.train.diff <- window(rain.diff, start = c(1,1), end = c(22,12))
rain.test.diff <- window(rain.diff, start = c(23,1))
```
```{r}
# Inspect that monthly seasonality was removed
autoplot(rain.train.diff[ ,2]) + theme_classic()

```

```{r}
# Train ETS model with differenced data
rain.ets <- ets(rain.train.diff[ , 2], model = 'ANN', alpha = 0.2)
rain.ets.pred <- forecast(rain.ets, h = length(rain.test), level = 0)
```
```{r}
accuracy(rain.ets.pred, rain.test[ , 2])
```
```{r}
autoplot(rain.train.diff[ , 2], series = "Training") +
  autolayer(rain.ets$fitted, series = 'Model', alpha = .8) +
  autolayer(rain.ets.pred, series = 'Prediction', alpha = .8) +
  autolayer(rain.test.diff[ , 2], series = "Test", alpha = .8) +
  theme_classic() +
  labs(y = 'Rain (mm)', title = 'Precipitation by Year') +
  coord_cartesian(xlim = c(20, 24)) +
  scale_x_continuous(breaks = c(20:24), labels = c(2020,2021,2022,2023,2024))

```

# ETS Model without Differencing - Performs Best on the Data
```{r}
rain.ets <- ets(rain.train[ , 2], model = 'ZNA')
rain.ets.pred <- forecast(rain.ets, h = length(rain.test), level = 0)
```
```{r}
accuracy(rain.ets.pred, rain.test[ , 2])
```
```{R}
autoplot(rain.train[ , 2], series = "Training") +
  autolayer(rain.ets$fitted, series = 'Model', alpha = .8) +
  autolayer(rain.ets.pred, series = 'Prediction', alpha = .8) +
  autolayer(rain.test[ , 2], series = "Test", alpha = .8) +
  theme_classic() +
  labs(y = 'Rain (mm)', title = 'Precipitation by Year') +
  coord_cartesian(xlim = c(20, 24)) +
  scale_x_continuous(breaks = c(20:24), labels = c(2020,2021,2022,2023,2024))
```
# Train Auto-Arima Models for Monthly Data
## Rain auto ARIMA: Model has decent metrics.
```{r}
#Auto model to start
rain.auto <- auto.arima(rain.train.y, xreg = rain.train.X)
summary(rain.auto)
```
# Train ARIMA Model
```{r}
rain.arima <- Arima(rain.train.y, order = c(0,1,1), seasonal = c(1,1,1), xreg = rain.train.X)
summary(rain.arima)
```

# Forecast with ARIMA model
```{r}
arima.pred <- forecast(rain.arima, xreg = rain.test.X)
```
```{r}
accuracy(arima.pred, rain.test.y)

```
```{r}
#Plot ARIMA Model
autoplot(rain.train.y, series = 'Training') +
  autolayer(rain.test.y, series = 'Test') +
  theme_classic() +
  labs(y = 'Rain (mm)', x = 'Month', title = 'Precipitation Forecast (12 Month)') +
  coord_cartesian(xlim = c(20, 24)) +
  scale_x_continuous(breaks = c(20:24), labels = c(2020,2021,2022,2023,2024))
```
```{r}
#Plot ARIMA Model
autoplot(rain.arima$fitted, series = 'Model') +
  autolayer(arima.pred, series = 'Prediction') +
  theme_classic() +
  labs(y = 'Rain (mm)', x = 'Month', title = 'Precipitation Forecast (12 Month)') +
  coord_cartesian(xlim = c(20, 24)) +
  scale_x_continuous(breaks = c(20:24), labels = c(2020,2021,2022,2023,2024))

```
# Plot Residuals against Real Values
```{r}
colnames(arima.df)[1] <- "point"
colnames(arima.df)[2] <- "low80"
colnames(arima.df)[3] <- "hi80"
colnames(arima.df)[4] <- "low95"
colnames(arima.df)[5] <- "hi95"

arima.df$actual <- rain.test.y                     

ggplot(arima.df, aes(x = actual, y = point)) +
  geom_point() +
  theme_classic()
```

# Logistic Regression for Season Question: What variables classify Rain Events above 0.5 inch/ 13 mm? 
# Training Rain as binary outcome and controlling for temperature swing and season
## Followed tutorial by Schmueli et al (2018) for cosine and sine variable 
```{r}
# ADd binary outcome to dataset as being rain over 13 mm (0.5 inch)
sd_rain_sn$bi <- if_else(sd_rain_sn$PRCP > 13, 1, 0)

# Combine temperature and rain data
sd_both <- merge(sd_rain_sn, sd_temp_sn, by = 'date')
#View(sd_both)

# Make outcome variable into a factor
sd_both$bi <- as.factor(sd_both$bi)

# Look at proportion of outcome variable
# Almost 50/50 split
prop.table(table(sd_both$bi))

# Make sine varibale for seasonality
sd_both$t <- seq(1, length(sd_both$bi), 1)
sd_both$sine <- sin(2 * pi * sd_both$t / 365.25)

# Make cosine varible for seasonality
sd_both$cosine <- cos(2 * pi * sd_both$t / 365.25)

# Add a Season Variable
season <- list('Winter', 'Spring', 'Summer', 'Fall')
sd_both$season <- rep(season, 83)
sd_both$season <- as.character(sd_both$season)
sd_both$season <- as.factor(sd_both$season)

# Get last 22 years of data
sd_both <- sd_both %>% filter(date > '1999-12-31')

# Inspect first few values to confirm
#head(sd_both)
```
```{r}
# Split data into testing and training
bi.train <- sd_both[c(1:68), ]
bi.test <- sd_both[c(68:92), ]
```

# Seasonal Max Temperature Difference and Cosine Maxmize prediction of target class 

```{r}
# Train logistic regression model
# Cosine predictor optimized ability to predict positive class
sd.log <- glm(bi ~ TDIFF + cosine + season, data = bi.train, family = 'binomial')
summary(sd.log)

```
```{r}
# predict testing period
log.pred <- predict(sd.log, bi.test[ , -3], type = 'response')
bi.test$pred <- if_else(log.pred > 0.5, 1, 0)
bi.test$pred <- as.factor(bi.test$pred)

# Create confusion matrix
confusionMatrix(bi.test$bi, bi.test$pred, positive = '1')

```


# Logistic Regression for Monthly Question: What variables classify Monthly Rain Events above 0.5 inch/ 13 mm? 
# Training Rain as binary outcome and controlling for temperature swing and season
## Followed tutorial by Schmueli et al (2018) for cosine and sine variable 
```{r}
View(sd_rain_mn)

# ADd binary outcome to dataset as being rain over 13 mm (0.5 of a inch)
sd_rain_mn$bi <- if_else(sd_rain_mn$PRCP > 13, 1, 0)

# Combine temperature and rain data
sd_both <- merge(sd_rain_mn, sd_temp_mn, by = 'date')
#View(sd_both)

# Make outcome variable into a factor
sd_both$bi <- as.factor(sd_both$bi)

# Look at proportion of outcome variable
# There is 28.9% rain events above 25 mm 
prop.table(table(sd_both$bi))

# Make sine varibale for seasonality
sd_both$t <- seq(1, length(sd_both$bi), 1)
sd_both$sine <- sin(2 * pi * sd_both$t / 365.25)

# Make cosine varible for seasonality
sd_both$cosine <- cos(2 * pi * sd_both$t / 365.25)

# Get last 22 years of data
sd_both <- sd_both %>% filter(date > '1999-12-31')

# Inspect first few values to confirm
View(sd_both)
```



```{r}
# Split data into testing and training
bi.train <- sd_both[c(1:264), ]
bi.test <- sd_both[c(265:276), ]
```

# Seasonal Max Temperature Difference and Cosine Maxmize prediction of target class 

```{r}
# Train logistic regression model
# Cosine predictor optimized ability to predict positive class
sd.log <- glm(bi ~ TDIFF, data = bi.train, family = 'binomial')
summary(sd.log)

```
```{r}
# predict testing period
log.pred <- predict(sd.log, bi.test[ , -3], type = 'response')
bi.test$pred <- if_else(log.pred > 0.5, 1, 0)
bi.test$pred <- as.factor(bi.test$pred)

# Create confusion matrix
confusionMatrix(bi.test$bi, bi.test$pred, positive = '1')

```

# References
## https://cran.r-project.org/web/packages/tibbletime/tibbletime.pdf
## schmueli